From physical conditions to Hamiltonian and Lagrangian mechanics

### 2021-01-04 Articles/2020-ClassicalMechanicsPhil/Source/classical-phil-article.tex
Update classical-phil-article.tex
Edits to parts of Section 4 (on new technical material, and fixed problems noted on 12-21) I still need to massively restructure this section, along with moving the discussion of the gauge redundancy of p out of section 5.2 and into this section.

* A paragraph to show how densities work. In general are coordinate dependent and you can't simply compare values at different points. This is the case in Lagrangian mechanics because both differentials in position vary controvariantly.

@@

@@

*[Could save the following for discussion of gauge redundancy in section 5.2] Not all ``state variables" are physical. Hamiltonian mechanics does not really care what the q's and p's are. It is the motion assumption that tells us that q's are position, simply because they define the unit system. P is a guage quantity (we care about the trajectory, the velocity). In this view, p is simply a ``label'' we can use to express the density over states. Therefore we do care about the physicality of the invariant distribution more than the physicality of the state variables themselves. -> Invariant density is really the important physical element/object in Hamiltonian mechanics.

@@

* Invariant densities: physical significance connected to coordinate independence. [incorporated]

@@

* Makes the fundamental role that phase space plays in statistical mechanics more physically salient/perspicuous. Areas/volumes are measures of number of states. They are coordinate invariant. Also the Shannon/Gibbs entropy are coordinate independent (see physicality of quantities that are coordinate invariant)

@@

We have now substantiated the claim of Section~\ref{privileging} that Hamiltonian mechanics requires less structure than Lagrangian mechanics, in virtue of requiring fewer physical conditions. While this provides one reason for privileging Hamiltonian mechanics, it is not the only one. A further---possibly even more important---reason stems from the relationship between Hamiltonian mechanics and invariant densities. As we argue in this section, viewing physics through the lens of invariant densities has much to commend it. In general, it provides a better understanding of classical systems. Specifically, we will see that invariant densities are of fundamental physical significance due to their coordinate-independence [invariance?]. Hamiltonian mechanics---unlike Lagrangian mechanics---allows us to compare these densities point-wise in any coordinate system. Although the density is physically significant because it is coordinate-invariant, we will nevertheless argue that coordinate-invariance is not a necessary condition for physical significance. In fact, the ability to express densities in some choice of units is closely connected with symplectic structure. Hence, we will argue that some coordinate-dependent representational choices are nevertheless physically significant. This view is at odds with metaphysically-driven demands to represent physical theories in exclusively ``intrinisic" terms [get citations: Field 1980; Chen papers; North]. 
% Section~\ref{action} uses the invariant density to provide a geometrical interpretation of the action principle.

In the Hamiltonian formalism, the density has the following desirable property: we can compare the density point-wise across phase space. If the scalar value of the density at one point is larger than another, this entails that the density is higher at that point of phase space. Physically, this means that there is more physical stuff at this point. Likewise, if the scalar value of the density is the same at two different points, this entails that there is the same amount of stuff at those points. For brevity, will refer to this as the property of \textit{point-wise comparison}. Since the density depends in general on the choice of units, point-wise comparison does not hold universally across representations of state space. Instead, for the density of states to have this property, we require a coordinate-independent definition of the density. As the derivation in Section~\ref{deterministic} shows, we must define our state space using a symplectic manifold in order to define an invariant density. In other words, to define a density with the property of point-wise comparison, we must use canonical coordinates $(q, p)$ defined on a symplectic manifold. As we will discuss further in Section~\ref{gauge}, the conjugate momentum $p$ is itself a gauge-dependent quantity. Hence, the real physical significance of canonical coordinates is that they allow us to define an invariant phase space density $\rho $.

@@

Section~\ref{fundamentality} analyzed fundamentality in terms of generality. We argued that Hamiltonian mechanics is more fundamental than Lagrangian mechanics in part because it is more general. By considering how Hamiltonian mechanics supports faithful point-wise density comparisons, we have seen two further notions of fundamentality. The first is epistemic: by representing the density on a symplectic manifold, we do not need to know which coordinate system we have chosen to work in. Any and all coordinate systems will support the formal property of point-wise comparisons [what about time-dependent coordinate transformations?]. Not only is this formal property methodologically advantageous, but it also bespeaks a further physical significance. This brings us to a second, physical notion of fundamentality: the phase space density $\rho$ is physically fundamental because it is a coordinate-invariant. Assuming that coordinate-invariance is sufficient for physical significance, the density thereby represents a genuine physical property of classical systems. It quantifies the amount of stuff in each phase space state. [to do: should we make this precise in terms of representational theory of measurement? E.g., the particular kind of scale that this representation defines and the meaningful comparisons it supports?].

% Here, we consider an epistemic sense of ``fundamental." In this sense, Hamiltonian mechanics is ``fundamental" because it shows that densities are coordinate-independent quantities. Assuming that coordinate-independence is sufficient for being physically significant, this shows that these densities are physically significant. More precisely, they are physical quantities or properties of the system.

>>>>

* A paragraph to show how densities work. In general are coordinate dependent and you can't simply compare values at different points. This is the case in Lagrangian mechanics because both differentials in position vary contravariantly.

@@

\subsection{Density and Riemmanian geometry}

It is helpful to take a quick look at how densities work in general so that we can better understand what is so special about them in the context in Hamiltonian mechanics. By density we mean the intuitive physical notion of a quantity divided by the region it occupies, like mass per volume or charge per area. Over a continuum, this is an integrable function $\rho(x)$. Mathematically, the concept is captured by the Radon–Nikodym derivative: we have a region $U$, we have two measures $m$ and $V$ and the derivative satistifies $m(U) = \int_U \rho dV$. This makes it clear that the units of $\rho$ depends on the units of both $m$ and $V$. More precisely, if $\rho(x, y)$ is a function of multiple variables, we will want $\int_U \rho(x, y) dx dy$ so that the units of $\rho$ are consistent with the units of the arguments. For example, in polar coordinates $\rho(r, \theta)$ will return mass over meter-radians if we measure $r$ in meters and $\theta$ in radians.

The density is coordinate dependent because its definition requires us to make a choice of coordinates. Worse than that, it is not even coordinate invariant: the density changes with the Jacobian. For example, changing from polar to Cartesian coordinates we have $\rho(x,y) = |J| \rho(r, \theta) = r \rho(r, \theta)$. Since the Jacobian is a function of the point, this makes the value of the density, in general, not directly comparable at different points. If we have a uniform density in $x$ and $y$, we will have an $r$-dependence in $r$ and $\theta$. This dependence is introduced precisely because the area identified by $dr$ and $d\theta$ is $dr \,r d\theta$, which is a function of $r$.

In differential geometry, a different mathematical object is called the `density,' namely a volume form (i.e. an alternating function from $n$ vectors to a real number) that is unaffected by changes in orientation. Our physically significant density $\rho$ is actually the component of this volume form $\nu = \rho(x^i) dx^1 \wedge dx^2 \wedge ... \wedge dx^n$. Note that $\nu$ combines the density and the volume into a coordinate-free object. Considering the total mass in the region $U$, we have $m(U) = \int \nu$ so essentially $\nu = dm$ which is the infinitesimal mass contribution we are integrating. But this means the value returned by $\nu$ depends not only on how must stuff there is at a given point, but also on the size of the volume identified by the $n$ vectors given to the form. [Not immediately clear to me what the upshot is of this last point; e.g. how does this relate to $\nu$ being coordinate-free?]

In Riemannian geometry, we have a canonical volume form $dV = \epsilon \sqrt{|g_{ab}|} dx^1 \wedge dx^2 \wedge ... \wedge dx^n$ where $|g_{ab}|$ is the determinant of the metric tensor as expressed in local coordinates, and $\epsilon$ is $1$ or $-1$ depending on the orientation. In component notation, we can express the volume form as $\sqrt{|g_{ab}|}\epsilon_{a_1 ... a_n}$, where $\epsilon_{a_1 ... a_n}$ is the Levi-Civita symbol. We can then express any differential density $\nu = \hat{\rho} dV$ as the product of the volume density times a scalar function $\hat{\rho}$. At first glance, this expression misleadingly suggests that the density $\nu$ is coordinate-independent, but this is not the case. [double check that this claim applies to $\nu$ rather than $\hat{\rho} $. Is this in tension with saying earlier that $\nu$ is coordinate-free?] The quantity $|g_{ab}|$ depends on the choice of coordinates and is, in general, a function of the point. However, we can always find a set of coordinates that, at least locally, makes $|g_{ab}|$ constant. The determinant $|g_{ab}|$ is the Jacobian of the transformation to those coordinates [the rationale/point of this sentence was not clear to me]. For example, in Cartesian coordinates $|g_{ab}| = 1$ and the volume form $dV = dx \wedge dy$; in polar coordinates $|g_{ab}| = r^2$ and the volume form $dV = r dr \wedge d\theta$ [above we said that the Jacobian was $r$ rather than $r^2$]. This means that $\hat{\rho}$ is really expressing the density in a particularly convenient Cartesian-like coordinate system: one in which we can compare the density point-wise. We can also see this if we look at the units of $\hat{\rho}$, which are in terms of inverse volume as expressed in the privileged system.\footnote{This is similar to what happens with the concept of proper time: it is defined in the comoving frame, so it is coordinate dependent. Yet, the expression $d\tau = \frac{1}{c} \sqrt{g_{ab} dx^a dx^b}$ is coordinate invariant.}

We can now look at what happens for distributions over position and velocity. Note that the velocities transform in the same way (i.e. contravariantly) as differentials in position. Therefore $\rho(x^i, v^j) = \sqrt{|g_{ab}|} \sqrt{|g_{ab}|} \rho(\hat{x}^i, \hat{v}^j) = |g_{ab}| \rho(\hat{x}^i, \hat{v}^j)$ $dx^a$. The determinant of the metric tensor is therefore directly the volume of the tangent bundle [I didn't follow this point]. The relationship to the density as expressed in phase space is therefore the following:
\begin{equation}
	\rho(q^i, p_j) = |g_{ab}| \rho(x^i, v^j).
\end{equation}
A change of coordinates will induce a contravariant change in the velocity and a covariant change in the momentum. The left side will remain invariant. The two terms on the right side will change in the opposite way so that their product is invariant. Focusing on densities, then, brings about a physically-motivated way to understand the relationship between the cotangent and the tangent bundle, where the metric tensor plays a role in defining the geometry of the second in a similar way that the symplectic form defines the geometry of the first: it defines the count of states so that densities can be properly tracked.

@@

*[Could save the following for discussion of gauge redundancy in section 5.2] Not all ``state variables" are physical. Hamiltonian mechanics does not really care what the q's and p's are. It is the motion assumption that tells us that q's are position, simply because they define the unit system. P is a guage quantity (we care about the trajectory, the velocity). In this view, p is simply a ``label'' we can use to express the density over states. Therefore we do care about the physicality of the invariant distribution more than the physicality of the state variables themselves. $->$ Invariant density is really the important physical element/object in Hamiltonian mechanics.

@@

* Invariant densities: physical significance connected to coordinate invariance. [incorporated]

@@

* Makes the fundamental role that phase space plays in statistical mechanics more physically salient/perspicuous. Areas/volumes are measures of number of states. They are coordinate invariant. Also the Shannon/Gibbs entropy are coordinate invariant [but not coordinate independent?] (see physicality of quantities that are coordinate invariant)

@@

We have now substantiated the claim of Section~\ref{privileging} that Hamiltonian mechanics requires less structure than Lagrangian mechanics, in virtue of requiring fewer physical conditions. While this provides one reason for privileging Hamiltonian mechanics, it is not the only one. A further---possibly even more important---reason stems from the relationship between Hamiltonian mechanics and invariant densities. As we argue in this section, viewing physics through the lens of invariant densities has much to commend it. In general, it provides a better understanding of classical systems. Specifically, we will see that densities are of fundamental physical significance due to their coordinate-invariance. Hamiltonian mechanics---unlike Lagrangian mechanics---allows us to compare these densities point-wise in any coordinate system. Although the density is coordinate-invariant, it is not coordinate-independent (it cannot be given a coordinate-free definition). This is because expressing the density requires a choice of units, a fact that is closely connected with symplectic structure. Hence, we will argue that some coordinate-dependent representational choices are nevertheless physically significant. This view is at odds with metaphysically-driven demands to represent physical theories in exclusively ``intrinisic" terms [get citations: Field 1980; Chen papers; North]. 
% Section~\ref{action} uses the invariant density to provide a geometrical interpretation of the action principle.
% Remember: the density is coordinate-invariant but not coordinate-independent; it is actually coordinate-dependent (i.e. it does not admit of something like a coordinate-free definition)
% Also remember: there are quantities that are both coordinate-invariant and coordinate -dependent that are still physically significant, e.g. choices of q's and p's (?)
% Coordinate-invariant in general means that definition of the quantity involves referencing coordinates, but if you change the coordinates, then the value of the quantity remains the same. (E.g. transforms as a scalar). so densities are never coordinate-independent. defining a density requires a choice of units, so involves referencing units. 
%see 57.3 minutes of 12-21 recording: when we define the density originally, it is coordinate-free (rho as a function of the state space of the particles, S). But when we go to express it numerically, then it becomes coordinate-dependent. (I.e. when we express it over a manifold, then it needs to be a density for integration, and hence it becomes coordinate-dependent). 
% see 51 minutes of 12-21 recording: Densities (the coordinate-dependent kind) are physically significant partly because they tell you how much stuff is in each state

In the Hamiltonian formalism, the density has the following desirable property: we can compare the density point-wise across phase space. If the scalar value of the density at one point is larger than another, this entails that the density is higher at that point of phase space. Physically, this means that there is more physical stuff at this point. Likewise, if the scalar value of the density is the same at two different points, this entails that there is the same amount of stuff at those points. For brevity, will refer to this as the property of \textit{point-wise comparison}. Since in general the density is coordinate-dependent, this property of point-wise comparison does not always hold. For instance, it can fail as we change from Cartesian to polar coordinates [figure out where to insert or reference the illustration of this shown above]. For point-wise comparison to hold, we require a coordinate-invariant density. As the derivation in Section~\ref{deterministic} shows, we must define our state space using a symplectic manifold in order to define an invariant density. In other words, to define a density with the property of point-wise comparison, we must use canonical coordinates $(q, p)$ defined on a symplectic manifold. As we will discuss further in Section~\ref{gauge}, the conjugate momentum $p$ is itself a gauge-dependent quantity. Hence, the real physical significance of canonical coordinates is that they allow us to define an invariant phase space density $\rho $. [Move this section 5.2 stuff up into this section; and figure out how to reorganize this paragraph]

@@

Section~\ref{fundamentality} analyzed fundamentality in terms of generality. We argued that Hamiltonian mechanics is more fundamental than Lagrangian mechanics in part because it is more general. By considering how Hamiltonian mechanics supports faithful point-wise density comparisons, we have seen two further notions of fundamentality. The first is epistemic: by representing the density on a symplectic manifold, we do not need to know which coordinate system we have chosen to work in. Any and all (time-independent) coordinate systems will support the formal property of point-wise comparisons.\footnote{Time-dependent coordinate transformations require working in the extended phase space. See~\ref{gauge} for discussion.} Not only is this formal property methodologically advantageous, but it also bespeaks a further physical significance. This brings us to a second, physical notion of fundamentality: the phase space density $\rho$ is physically fundamental because it tracks how much of the system is in each possible state. It quantifies the amount of stuff in each phase space state. Section~\ref{deterministic} demonstrated what is perhaps a third reason for viewing the density as fundamental: by defining the density, we recover the right mathematical structure, namely a symplectic manifold. If we did not define the density, then we would need another explanation of why this mathematical structure arises for classical systems. [I think this final point might be repetitive given a claim made in the next paragraph]
%see 11:50am of 12-21 meeting video for clarification of this point. There are probably finer details for me to incorporate.

% We took out the following because coordinate-invariance in fact did not seem to be either a necessary or sufficient condition for physical significance: Assuming that coordinate-invariance is sufficient for physical significance, the density thereby represents a genuine physical property of classical systems.
